---
title: "Correlation Analysis"
author: "Mari-Lee Odendaal"
date: "`r Sys.time()`"
---

## Load R-packages

```{r}
library(here);library(glue);library(colorspace);library(pheatmap);library(stringr)
```

```{r setup, include=F, message=F}
subdir_name <- "correlation_analysis"

# set paths
knitr::opts_knit$set(root.dir=".", aliases=c(h = "fig.height", w = "fig.width", ow = "out.width"))
knitr::opts_chunk$set(dev=c('png', 'pdf'), 
                      fig.path=here("results", "figures", glue("{subdir_name}/")),
                      dpi=300)
```

```{r knit, echo=F, eval=FALSE}
rmarkdown::render(input = here("scripts", str_c(subdir_name, ".Rmd")), output_dir = here("results"))
```

## Input R-objects

```{r}
cpm_adults <- readRDS("cpm_adults.RDS")
cpm_juveniles <- readRDS("cpm_juveniles.RDS")
```

## Pre-processing for the correlation analyses

```{r}
colnames(cpm_adults) <- c("Future S1","Future S2","Past S1","Past S2","Present S1","Present S2")
cpm_adults <- cpm_adults[,c(3,4,5,6,1,2)]
m_ad <- data.matrix(cpm_adults)
```

Create a data matrix for the adults.

```{r}
colnames(cpm_juveniles) <- c("Future S1","Future S2","Past S1","Past S2","Present S1","Present S2")
cpm_juveniles<- cpm_juveniles[,c(3,4,5,6,1,2)]
m_juv <- data.matrix(cpm_juveniles)
```

Create a data matrix for the juveniles.

```{r}
ri_0_1_2 <- c(0,0,1,1,2,2)
ri_0_1_0 <- c(0,0,1,1,0,0)
ri_1_0_0 <- c(1,1,0,0,0,0)
ri_0_0_1 <- c(0,0,0,0,1,1)
```

Make variables for the four different expression patterns.

```{r}
paletteLength <- 15
col <- diverging_hcl(15, "Red-Green")
col <- rev(col)
```

Set the colors for visualization.

```{r}
cpm_adults <- as.data.frame(cpm_adults)
cpm_adults$gene <- rownames(cpm_adults)
cpm_juveniles <- as.data.frame(cpm_juveniles)
cpm_juveniles$gene <- rownames(cpm_juveniles)
```

## Pearson correlation analyses for the adults
### Genes correlated with decreasing pH

```{r results=F}
res <- apply(m_ad,1L,cor.test,ri_0_1_2)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_ad)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_adults_0_1_2 <- as.data.frame(merge(cpm_adults, pvalues, by = "gene", all = T))
rownames(cpm_adults_0_1_2) <- cpm_adults_0_1_2$gene
cpm_adults_0_1_2 <- cpm_adults_0_1_2[cpm_adults_0_1_2$pvalues < 0.05,]

cpm_adults_0_1_2 <- cpm_adults_0_1_2[,2:7]
cpm_adults_0_1_2 <- t(scale(t(cpm_adults_0_1_2)))
```

4.5% of the genes is correlated with decreasing pH.

```{r heatmap_adult_0_1_2, ow = '90%', h = 3, w = 5}
hm_ad_0_1_2 <- pheatmap(cpm_adults_0_1_2, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

### Genes correlated with change in ambient environment

```{r results=F}
res <- apply(m_ad,1L,cor.test,ri_0_1_0)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_ad)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_adults_0_1_0 <- as.data.frame(merge(cpm_adults, pvalues, by = "gene", all = T))
rownames(cpm_adults_0_1_0) <- cpm_adults_0_1_0$gene
cpm_adults_0_1_0 <- cpm_adults_0_1_0[cpm_adults_0_1_0$pvalues < 0.05,]

cpm_adults_0_1_0 <- cpm_adults_0_1_0[,2:7]
cpm_adults_0_1_0 <- t(scale(t(cpm_adults_0_1_0)))
```

8.6% of the genes is correlated with change in the ambient environment.

```{r heatmap_adult_0_1_0, ow = '90%', h = 5, w = 5}
hm_ad_0_1_0 <- pheatmap(cpm_adults_0_1_0, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

### Genes with a strong (low/high) expression in the past condition

```{r results=F}
res <- apply(m_ad,1L,cor.test,ri_1_0_0)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_ad)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_adults_1_0_0<- as.data.frame(merge(cpm_adults, pvalues, by = "gene", all = T))
rownames(cpm_adults_1_0_0) <- cpm_adults_1_0_0$gene
cpm_adults_1_0_0 <- cpm_adults_1_0_0[cpm_adults_1_0_0$pvalues < 0.05,]

cpm_adults_1_0_0 <- cpm_adults_1_0_0[,2:7]
cpm_adults_1_0_0 <- t(scale(t(cpm_adults_1_0_0)))
```

4.0% of the genes has a strong (low/high) expression in the past condition.

```{r heatmap_adult_1_0_0, ow = '90%', h = 3, w = 5}
hm_ad_1_0_0 <- pheatmap(cpm_adults_1_0_0, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

### Genes with a strong (low/high) expression in the future condition

```{r results=F}
res <- apply(m_ad,1L,cor.test,ri_0_0_1)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_ad)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_adults_0_0_1<- as.data.frame(merge(cpm_adults, pvalues, by = "gene", all = T))
rownames(cpm_adults_0_0_1) <- cpm_adults_0_0_1$gene
cpm_adults_0_0_1 <- cpm_adults_0_0_1[cpm_adults_0_0_1$pvalues < 0.05,]

cpm_adults_0_0_1 <- cpm_adults_0_0_1[,2:7]
cpm_adults_0_0_1 <- t(scale(t(cpm_adults_0_0_1)))
```

4.6% of the genes has a strong (low/high) expression in the future condition.

```{r heatmap_adult_0_0_1, ow = '90%', h = 3, w = 5}
hm_ad_0_0_1<- pheatmap(cpm_adults_0_0_1, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

## Pearson correlation analyses for the juveniles
### Genes correlated with decreasing pH

```{r results=F}
res <- apply(m_juv,1L,cor.test,ri_0_1_2)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_juv)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_juv_0_1_2 <- as.data.frame(merge(cpm_juveniles, pvalues, by = "gene", all = T))
rownames(cpm_juv_0_1_2) <- cpm_juv_0_1_2$gene
cpm_juv_0_1_2 <- cpm_juv_0_1_2[cpm_juv_0_1_2$pvalues < 0.05,]

cpm_juv_0_1_2 <- cpm_juv_0_1_2[,2:7]
cpm_juv_0_1_2 <- t(scale(t(cpm_juv_0_1_2)))
```

7.7% of the genes is correlated with decreasing pH.

```{r heatmap_juv_0_1_2, ow = '90%', h = 5, w = 5}
hm_juv_0_1_2 <- pheatmap(cpm_juv_0_1_2, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

### Genes correlated with change in ambient environment

```{r results=F}
res <- apply(m_juv,1L,cor.test,ri_0_1_0)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_juv)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_juv_0_1_0 <- as.data.frame(merge(cpm_juveniles, pvalues, by = "gene", all = T))
rownames(cpm_juv_0_1_0) <- cpm_juv_0_1_0$gene
cpm_juv_0_1_0 <- cpm_juv_0_1_0[cpm_juv_0_1_0$pvalues < 0.05,]

cpm_juv_0_1_0 <- cpm_juv_0_1_0[,2:7]
cpm_juv_0_1_0 <- t(scale(t(cpm_juv_0_1_0)))
```

3.9% of the genes is correlated with change in the ambient environment.

```{r heatmap_juv_0_1_0, ow = '90%', h = 3, w = 5}
hm_juv_0_1_0 <- pheatmap(cpm_juv_0_1_0, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

### Genes with a strong (low/high) expression in the past condition

```{r results=F}
res <- apply(m_juv,1L,cor.test,ri_1_0_0)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_juv)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_juv_1_0_0 <- as.data.frame(merge(cpm_juveniles, pvalues, by = "gene", all = T))
rownames(cpm_juv_1_0_0) <- cpm_juv_1_0_0$gene
cpm_juv_1_0_0 <- cpm_juv_1_0_0[cpm_juv_1_0_0$pvalues < 0.05,]

cpm_juv_1_0_0 <- cpm_juv_1_0_0[,2:7]
cpm_juv_1_0_0 <- t(scale(t(cpm_juv_1_0_0)))
```

10.2% of the genes has a strong (low/high) expression in the past condition.

```{r heatmap_juv_1_0_0, ow = '90%', h = 6, w = 5}
hm_juv_1_0_0 <- pheatmap(cpm_juv_1_0_0, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

### Genes with a strong (low/high) expression in the future condition

```{r results=F}
res <- apply(m_juv,1L,cor.test,ri_0_0_1)

pvalues <- sapply(res,'[[','p.value')
pvalues <- as.data.frame(pvalues)
table(pvalues$pvalues < 0.05)[2]/dim(m_juv)[1]*100
pvalues$gene <- rownames(pvalues)

cpm_juv_0_0_1 <- as.data.frame(merge(cpm_juveniles, pvalues, by = "gene", all = T))
rownames(cpm_juv_0_0_1) <- cpm_juv_0_0_1$gene
cpm_juv_0_0_1 <- cpm_juv_0_0_1[cpm_juv_0_0_1$pvalues < 0.05,]

cpm_juv_0_0_1 <- cpm_juv_0_0_1[,2:7]
cpm_juv_0_0_1 <- t(scale(t(cpm_juv_0_0_1)))
```

4.1% of the genes has a strong (low/high) expression in the future condition.

```{r heatmap_juv_0_0_1, ow = '90%', h = 3, w = 5}
hm_juv_0_0_1 <- pheatmap(cpm_juv_0_0_1, col=col, show_rownames = F, cluster_cols = F, cellwidth = 20, cellheight = 0.2)
```

## Session information

```{r}
sessionInfo()
```

