---
title: "GO enrichment analysis"
author: "Mari-Lee Odendaal"
date: "`r Sys.time()`"
params: 
      subdir_name: enrichment_analysis
---

## Load R-packages

```{r message=F}
library(pheatmap);library(tidyverse);library(tidyverse);library(here);library(glue)
```

```{r setup, include=F, message=F}
subdir_name <- params$subdir_name

# set paths
knitr::opts_knit$set(root.dir=".", aliases=c(h = "fig.height", w = "fig.width", ow = "out.width"))
knitr::opts_chunk$set(dev=c('png', 'pdf'), 
                      fig.path=here("results", "figures", glue("{subdir_name}/")),
                      dpi=300)
```

```{r knit, echo=F, eval=FALSE}
rmarkdown::render(input = here("scripts", str_c(subdir_name, ".Rmd")), output_dir = here("results"))
```

## Input files

```{r}
fvsp_MF_adult <- readRDS("fvsp_MF_adult.RDS")
fvsp_MF_juvenile <- readRDS("fvsp_MF_juvenile.RDS")
fvsp_BP_adult <- readRDS("fvsp_BP_adult.RDS")
fvsp_BP_juvenile <- readRDS("fvsp_BP_juvenile.RDS")
pvsp_MF_adult <- readRDS("pvsp_MF_adult.RDS")
pvsp_MF_juvenile <- readRDS("pvsp_MF_juvenile.RDS")
pvsp_BP_adult <- readRDS("pvsp_BP_adult.RDS")
pvsp_BP_juvenile <- readRDS("pvsp_BP_juvenile.RDS")
```

Enriched GO terms associated with molecular function and biological process are determined by the GO MWU analysis (https://github.com/z0on/GO_MWU). The analysis is performed for each of the two comparisons in adult and juvenile.

```{r}
dir = here::here("input")
MF_ann <- read.table(file.path(dir, "GO_annotation_MF.txt"), sep = "\t", header = T)
BP_ann <- read.table(file.path(dir, "GO_annotation_BP.txt"), sep = "\t", header = T)
```

Import the grouping information

## Pre-processing for visualization of the enriched GO terms

```{r}
fvsp_MF_adult <- fvsp_MF_adult %>% mutate(group = "fvsp_MF_adult",
                                                GO = rownames(.))
fvsp_MF_juvenile <- fvsp_MF_juvenile %>% mutate(group = "fvsp_MF_juvenile",
                                                GO = rownames(.))
fvsp_BP_adult <- fvsp_BP_adult %>% mutate(group = "fvsp_BP_adult",
                                                GO = rownames(.))
fvsp_BP_juvenile <- fvsp_BP_juvenile %>% mutate(group = "fvsp_BP_juvenile",
                                                GO = rownames(.))
pvsp_MF_adult <- pvsp_MF_adult %>% mutate(group = "pvsp_MF_adult",
                                                GO = rownames(.))
pvsp_MF_juvenile <- pvsp_MF_juvenile %>% mutate(group = "pvsp_MF_juvenile",
                                                GO = rownames(.))
pvsp_BP_adult <- pvsp_BP_adult %>% mutate(group = "pvsp_BP_adult",
                                                GO = rownames(.))
pvsp_BP_juvenile <- pvsp_BP_juvenile %>% mutate(group = "pvsp_BP_juvenile",
                                                GO = rownames(.))
```

```{r}
df_MF <- rbind(pvsp_MF_adult, fvsp_MF_adult, pvsp_MF_juvenile, fvsp_MF_juvenile)
df_BP <- rbind(pvsp_BP_adult, fvsp_BP_adult, pvsp_BP_juvenile, fvsp_BP_juvenile)

df_BP <- df_BP[df_BP$pval < 0.05,]
df_MF <- df_MF[df_MF$pval < 0.05,]
```

GO terms that have a P-value of less than 0.05 are removed from the data.

```{r}
df_BP$GO <- sub(".*? ", "", df_BP$GO)
df_MF$GO <- sub(".*? ", "", df_MF$GO)

df_MF <- df_MF %>% mutate(direction = ifelse(direction == 0, -1, direction),
                          pval = pval*direction)
df_BP <- df_BP %>% mutate(direction = ifelse(direction == 0, -1, direction),
                          pval = pval*direction)
```

```{r}
df_BP2 <- subset(df_BP, select = c(GO, group, pval)) %>% pivot_wider(names_from = group, values_from = pval) %>% as.data.frame()
rownames(df_BP2) <- df_BP2$GO
df_BP2 <- df_BP2[2:5]
df_BP2[is.na(df_BP2)] <- 0

keep <- rowSums(df_BP2 != 0) >= 2
df_BP2 <- df_BP2[keep,]
rownames(df_BP2) <- ifelse(rownames(df_BP2) %in% BP_ann$GO, BP_ann$GO2, NA)
```

Create the data frame of GO terms that are associated with biological process.

```{r}
df_MF2 <- subset(df_MF, select = c(GO, group, pval)) %>% pivot_wider(names_from = group, values_from = pval) %>% as.data.frame()
rownames(df_MF2) <- df_MF2$GO
df_MF2 <- df_MF2[2:5]
df_MF2[is.na(df_MF2)] <- 0

keep <- rowSums(df_MF2 != 0) >= 2
df_MF2 <- df_MF2[keep,]
rownames(df_MF2) <- ifelse(rownames(df_MF2) %in% MF_ann$GO, MF_ann$GO2, NA)
```

Create the data frame of GO terms that are associated with molecular function.

GO terms that are significantly enriched in at least 2 of the comparisons are kept in the data for visualization.

## Visualization of the enriched GO terms

```{r}
col <- c("#B6D7E8", "#4E9AC6", "#195696","#F7F7F7", "#9C1127", "#DA6954", "#F9C3A9")
breaks = c(-0.05, -0.01, -0.001, -1e-20, 1e-20, 0.001, 0.01, 0.05)
ann_col_MF <- c("#CAB2D6", "#6A3D9A", "#FB9A99", "#E31A1C")
ann_col_BP <- c("#1F78B4", "#A6CEE3", "#33A02C", "#B2DF8A", "#FFFF99")

ann_col <- list(MF = ann_col_MF,
                BP = ann_col_BP)
names(ann_col$MF) <- unique(MF_ann$MF)
names(ann_col$BP) <- unique(BP_ann$BP)
MF <- MF_ann
rownames(MF) <- MF_ann$GO2
MF <- subset(MF, select = MF)

BP <- BP_ann
rownames(BP) <- BP_ann$GO2
BP <- subset(BP, select = BP)
```

Set the colors and breaks for the heatmap and the group annotation.

```{r GO_MF, ow = '90%', h = 4, w = 11}
hm_MF <- pheatmap(df_MF2, col=col, show_rownames = T, cluster_cols = F,
breaks = breaks, cluster_rows = T, cellwidth = 20, cellheight = 12, annotation_row = MF, annotation_colors = ann_col, legend = F)
```

```{r GO_BP, ow = '90%', h = 4, w = 11}
hm_BP <- pheatmap(df_BP2, col=col, show_rownames = T, cluster_cols = F,
breaks = breaks, cluster_rows = T, cellwidth = 20, cellheight = 12, annotation_row = BP, annotation_colors = ann_col, legend = F)
```

## Session information

```{r}
sessionInfo()
```

